### 2019-05-27 Thesis/games.tex
Nightly build.
Monthly commit.

\chapter{A Fully Abstract Game Semantics for Idealized Algol}

@@

@@

It is also easy to see from Proposition \ref{PropCopycat} that the identity function $\id\from M_A \to M_A$ is a structural isomorphism from $A$ to itself, and that the resulting copycat strategy $\cc_{\id}$ is an identity for composition.  
Combining this with our result for associativity in the previous chapter, we get that
\begin{theorem}
  The collection of games forms a category $\G$, where the morphisms $A \to B$ are strategies for $A\implies B$, composition is as above and the identity morphisms are the copycat strategies induced from the identity functions on moves.
\end{theorem}

@@

We can generalize the subset inclusions of the previous chapter to \emph{tree immersions}.  
Tree immersions are similar to subset inclusions, but generated by a function between plays, rather than between moves.  
A consequence of this is that while tree immersions do give rise to strategies, these strategies are not in general innocent.

@@

We have now given the category-theoretic properties of all the connectives from chapter \ref{SecConnectives}, with the exception of the sequoid $\sequoid$ and the exponential $\oc$.

@@

>>>>

\chapter{A Sequoidal Game Semantics for Idealized Algol}
\label{ChapGames}

@@

The types of Idealized Algol are then defined by the following inductive grammar.

\[
  T \Coloneqq \com \mid \bool \mid \nat \mid \Var \mid T \to T\,.
  \]

@@

It is also easy to see from Proposition \ref{PropCopycat} that the identity function $\id\from M_A \to M_A$ is a structural isomorphism from $A$ to itself, and that the resulting copycat strategy $\cc_{\id}$ is an identity for composition.  
Combining this with our result for associativity in the previous section, we get that
\begin{theorem}
  The collection of games forms a category $\G$, where the morphisms $A \to B$ are strategies for $A\implies B$, composition is as above and the identity morphisms are the copycat strategies induced from the identity functions on moves.
\end{theorem}

@@

We can generalize the subset inclusions of the previous section to \emph{tree immersions}.  
Tree immersions are similar to subset inclusions, but generated by a function between plays, rather than between moves.  
A consequence of this is that while tree immersions do give rise to strategies, these strategies are not in general innocent.

@@

We have now given the category-theoretic properties of all the connectives from section \ref{SecConnectives}, with the exception of the sequoid $\sequoid$ and the exponential $\oc$.

@@

\chapter{Operational Semantics, Computational Adequacy and Full Abstraction}
\label{ChapFullAbstraction}

### 2019-05-27 Thesis/pmonads2.tex
Monthly commit.

\chapter{Parametric Monads and Full Abstraction}

@@

\begin{definition}
  Let $\sigma,\sigma'\from A \to B$ be morphisms in $\G/\oppcat{(\Rv_{\Omega}^{FS})}$, where $\G$ is the category of arenas and single-threaded strategies.  
  We say that $\sigma\approx_{\bP}\sigma'$ if for all legal plays $s$ of $A \to B$ we have
  \[
    \bP(s\in\sigma) = \bP(s\in\sigma')\,.
    \]
\end{definition}

@@

\begin{proposition}
  Let $\sigma\from A \to B$, $\tau\from B \to C$ be morphisms in $\G/\oppcat{(\Rv_{\Omega}^{FS})}$.  
  Let $s$ be a legal play in $A \to C$.
  Then
  \[
    \bP(s\in\sigma;\tau) = \sum_{\s\in\wit_B(s)}\bP(\s\vert_{A,B}\in\sigma)\bP(\s\vert_{B,C}\in\tau)\,.
    \]
\end{proposition}
\begin{proof}
  Suppose that $\sigma$ and $\tau$ are given by (equivalence classes of) pairs
  \begin{mathpar}
    (V\from\Omega\to X, \tilde \sigma \from A \to (\ul X \to B))
    \and
    (W\from\Omega\to Y, \tilde \tau \from B \to (\ul Y \to C))\,,
  \end{mathpar}
  where $V$ and $W$ are random variables and $\tilde \sigma,\tilde\tau$ are strategies in $\G$.
  Then the composition $\sigma;\tau$ in $\G/\oppcat{(\Rv_{\Omega}^{FS})}$ is given by $V\tensor W\from\Omega \to X\times Y$, together with the \Mellies composition
  \[
    A \xrightarrow{\tilde\sigma}
    (\ul{X} \to B) \xrightarrow{\ul{X} \to \tilde\tau}
    (\ul X \to (\ul Y \to C)) \to
    ((\ul X \times \ul Y) \to C) \xrightarrow{\mu\to C}
    (\ul{X\times Y} \to C)\,,
    \]
  where $\mu$ is $\langle\ul{\pr_X},\ul{\pr_Y}\rangle$.

  First suppose that $s$ is a sequence in $A \to C$, and let $\s\in\wit_B(s)$.  
  Let $\t$ be a sequence in $\sigma\|(\ul X \to \tau)$ such that $\t\vert_{A,B,C}=\s$.  
  Then $\t\vert_{A,\ul X \to B}\in\sigma$ and $\t\vert_{B,\ul Y,C}\in\tau$.

  Moreover, since we have $\t\vert_{A,B}=\s\vert_{A,B}$ and $\t\vert_{B,C}=\s\vert_{B,C}$, we must have
  \begin{mathpar}
    \t\vert_{\ul X} \in \Acc_{\s\vert_{A,B}}(\sigma)
    \and
    \t\vert_{\ul Y} \in \Acc_{\s\vert_{B,C}}(\tau)\,,
  \end{mathpar}
  where we have identified a play $qu$ occurring in the arena $\ul X$ with its underlying sequence $u$ of elements of $X$, and likewise for $Y$.

  This gives us a function
  \[
    \{\t\in\sigma\|(\ul X \to \tau)\suchthat \t\vert_{A,B,C}=\s\} \to \Acc_{\s\vert_{A,B}}(\sigma)\times\Acc_{\s\vert_{B,C}}(\tau)\,.
    \]
  We claim that this function is a bijection.

  Indeed, suppose that $\t,\t'$ are two interactions in $\sigma\|(\ul X \to \tau)$ such that $\t\vert_{A,B,C}=\t'\vert_{A,B,C}=\s$, $\t\vert_{\ul X} = \t'\vert_{\ul X}$ and $\t\vert_{\ul Y}=\t'\vert_{\ul Y}$.
  We claim that $\t = \t'$.

@@

  By our earlier analysis (see, for example, the proof of \ref{PropComposition}), we know that $p$ and $q$ must either both occur in the $A \to (\ul X \to B)$-component, or both in the $(\ul X \to B) \to (\ul X \to (\ul Y \to C))$-component.
  But since $\t,\t'\in\sigma\|(\ul X \to \tau)$ are both interactions of deterministic strategies, we also know that they must both be $O$-moves in that component -- otherwise, they would have to be equal.
  In particular, neither $p$ nor $q$ may be a move in the middle component $\ul X \to B$, since then it would be a $P$-move in one of the two components.

  Therefore, $p$ and $q$ are both $O$-moves in one of the outer components $A$ and $\ul X \to (\ul Y \to C)$.  
  By Corollary \ref{CorSwitchingCondition}, we know that only Player $P$ may switch between games in $\ul X \to (\ul Y \to C)$, and therefore $p$ and $q$ must occur in the same component game -- i.e., both in $A$, both in $\ul X$, both in $\ul Y$ or both in $C$.  
  But now the conditions that $\t\vert_{A,B,C}=\t'\vert_{A,B,C}$, $\t\vert_{\ul X}=\t'\vert_{\ul X}$ and $\t\vert_{\ul Y}=\t'\vert_{\ul Y}$ mean that we must have $p=q$, which is the desired contradiction.

  For surjectivity, let $u\in\Acc_{\s\vert_{A,B}}(\sigma)$ and $v\in\Acc_{\s\vert_{B,C}}(\tau)$.  
  We seek a sequence $\t\in\sigma\|(\ul X \to \tau)$ such that $\t\vert_{A,B,C}=\s$, $\t\vert_{\ul X}=qu$ and $\t\vert_{\ul Y}=qv$.
\end{proof}

\section{Danos-Harmer Probabilistic Game Semantics}

>>>>

\chapter{Parametric Monads and Full Abstraction}
\label{ChapParametricMonadsFullAbstraction}

@@

\begin{definition}
  Let $\sigma,\sigma'\from A \to B$ be morphisms in $\G/\oppcat{(\Rv_{\Omega}^{FS})}$, where $\G$ is the category of arenas and single-threaded strategies.  
  We say that $\sigma\approx_{\bP}\sigma'$ if for all legal plays $s$ of $A \to B$ we have
  \[
    \bP(s\in\sigma) = \bP(s\in\sigma')\,.
    \]
  \label{DefProbabilisticEquivalenceRelation}
\end{definition}

@@

\begin{proposition}
  Let $\sigma\from A \to B$, $\tau\from B \to C$ be morphisms in $\G/\oppcat{(\Rv_{\Omega}^{FS})}$.  
  Let $s$ be a legal play in $A \to C$.
  Then
  \[
    \bP(s\in\sigma;\tau) = \sum_{\s\in\wit_B(s)}\bP(\s\vert_{A,B}\in\sigma)\bP(\s\vert_{B,C}\in\tau)\,.
    \]
  \label{PropCompositionProb}
\end{proposition}
\begin{proof}
  Suppose that $\sigma$ and $\tau$ are given by (equivalence classes of) pairs
  \begin{mathpar}
    (V\from\Omega\to X, \tilde \sigma \from A \to (\ul X \to B))
    \and
    (W\from\Omega\to Y, \tilde \tau \from B \to (\ul Y \to C))\,,
  \end{mathpar}
  where $V$ and $W$ are random variables and $\tilde \sigma,\tilde\tau$ are strategies in $\G$.
  Then the composition $\sigma;\tau$ in $\G/\oppcat{(\Rv_{\Omega}^{FS})}$ is given by $V\tensor W\from\Omega \to X\times Y$, together with the \Mellies composition
  \[
    A \xrightarrow{\tilde\sigma}
    (\ul{X} \to B) \xrightarrow{\ul{X} \to \tilde\tau}
    (\ul X \to (\ul Y \to C)) \to
    ((\ul X \times \ul Y) \to C) \xrightarrow{\mu\to C}
    (\ul{X\times Y} \to C)\,,
    \]
  where $\mu$ is $\langle\ul{\pr_X},\ul{\pr_Y}\rangle$.

  First suppose that $s$ is a sequence in $A \to C$, and let $\s\in\wit_B(s)$.  
  Let $\t$ be a sequence in $\tilde\sigma\|(\ul X \to \tau)$ such that $\t\vert_{A,B,C}=\s$.  
  Then $\t\vert_{A,\ul X \to B}\in\tilde\sigma$ and $\t\vert_{B,\ul Y,C}\in\tilde\tau$.

  Moreover, since we have $\t\vert_{A,B}=\s\vert_{A,B}$ and $\t\vert_{B,C}=\s\vert_{B,C}$, we must have
  \begin{mathpar}
    \t\vert_{\ul X} \in \Acc_{\s\vert_{A,B}}(\tilde\sigma)
    \and
    \t\vert_{\ul Y} \in \Acc_{\s\vert_{B,C}}(\tilde\tau)\,,
  \end{mathpar}
  where we have identified a play $qu$ occurring in the arena $\ul X$ with its underlying sequence $u$ of elements of $X$, and likewise for $Y$.

  This gives us a function
  \[
    \{\t\in\tilde\sigma\|(\ul X \to \tilde\tau)\suchthat \t\vert_{A,B,C}=\s\} \to \Acc_{\s\vert_{A,B}}(\tilde\sigma)\times\Acc_{\s\vert_{B,C}}(\tilde\tau)\,.
    \]
  We claim that this function is a bijection.

  Indeed, suppose that $\t,\t'$ are two interactions in $\tilde\sigma\|(\ul X \to \tilde\tau)$ such that $\t\vert_{A,B,C}=\t'\vert_{A,B,C}=\s$, $\t\vert_{\ul X} = \t'\vert_{\ul X}$ and $\t\vert_{\ul Y}=\t'\vert_{\ul Y}$.
  We claim that $\t = \t'$.

@@

  By our earlier analysis (see, for example, the proof of \ref{PropComposition}), we know that $p$ and $q$ must either both occur in the $A \to (\ul X \to B)$-component, or both in the $(\ul X \to B) \to (\ul X \to (\ul Y \to C))$-component.
  But since $\t,\t'\in\tilde\sigma\|(\ul X \to \tilde\tau)$ are both interactions of deterministic strategies, we also know that they must both be $O$-moves in that component -- otherwise, they would have to be equal.
  In particular, neither $p$ nor $q$ may be a move in the middle component $\ul X \to B$, since then it would be a $P$-move in one of the two components.

  Therefore, $p$ and $q$ are both $O$-moves in one of the outer components $A$ and $\ul X \to (\ul Y \to C)$.  
  By Corollary \ref{CorSwitchingCondition}, we know that only Player $P$ may switch between games in $\ul X \to (\ul Y \to C)$, and therefore $p$ and $q$ must occur in the same component game -- i.e., both in $A$, both in $\ul X$, both in $\ul Y$ or both in $C$.  
  But now the conditions that $\t\vert_{A,B,C}=\t'\vert_{A,B,C}$, $\t\vert_{\ul X}=\t'\vert_{\ul X}$ and $\t\vert_{\ul Y}=\t'\vert_{\ul Y}$ mean that we must have $p=q$.  
  For example, if $p$ and $q$ are both moves in $C$, then we have $\r\vert_Cp\prefix\t\vert_C$ and $\r\vert_Cq\prefix \t'\vert_C$; since $\t\vert_C=\t'\vert_C$, we must have $p=q$.  
  This is the desired contradiction.

  For surjectivity, let $u\in\Acc_{\s\vert_{A,B}}(\sigma)$ and $v\in\Acc_{\s\vert_{B,C}}(\tau)$.  
  We seek a sequence $\t\in\tilde\sigma\|(\ul X \to \tilde\tau)$ such that $\t\vert_{A,B,C}=\s$, $\t\vert_{\ul X}=qu$ and $\t\vert_{\ul Y}=qv$.

  Since $u\in\Acc_{\s\vert_{A,B}}(\sigma)$, there is some sequence $s\in\sigma$ such that $s\vert_{A,B}=\s\vert_{A,B}$ and $s\vert_{\ul X}=qu$.  
  Similarly, since $v\in\Acc_{\s\vert_{B,C}}(\tau)$, there is some sequence $t\in\tau$ such that $t\vert_{B,C}=\s\vert_{B,C}$ and $t\vert_{\ul Y}=qv$.
  We form the sequence $\t$ as a suitable interleaving of $s$ and $t$, noting that they have the same $B$-components: we start with the sequence $\s$, and then insert the appropriate moves from (the left-hand copy of) $\ul X$ and $\ul Y$ between the corresponding moves from $A$, $B$ and $C$.  
  Lastly, we insert moves from the right-hand copy of $\ul X$ adjacent to the corresponding moves in the right-hand copy, inserting an $O$-move $q$ in the right-hand copy of $\ul X$ immediately after each $O$-move $q$ in the left-hand copy, and a $P$-move $x$ in the right-hand copy immediately before each $P$-move $x$ in the left-hand copy.

  This tells us that we have
{\setlength{\IEEEnormaljot}{20pt}%
  \begin{IEEEeqnarray*}{Cl}
    &  \bP(\s\vert_{A,B}\in\sigma)\bP(\s\vert_{B,C}\in\tau) \\
    = & \left( \sum_{u\in\Acc_{\tilde\sigma}(\s\vert_{A,B})} \prod_{i=0}^{|u|-1} \bP(V = u^{(i)}) \right)
        \left( \sum_{v\in\Acc_{\tilde\tau}  (\s\vert_{A,B})} \prod_{i=0}^{|v|-1} \bP(W = v^{(i)}) \right) \\
    = & \sum_{(u,v)\in\Acc_{\tilde\sigma}(\s\vert_{A,B}) \times \Acc_{\tilde\tau}(\s\vert_{B,C})} \prod_{i=0}^{|u|-1} \bP(V=u^{(i)}) \prod_{i=1}^{|v|-1} \bP(W=v^{(i)}) \\
    = & \sum_{\stackrel{\t\in\tilde\sigma\|(\ul X \to \tilde\tau)}{\t\vert_{A,B,C}=\s}}
      \left[ \prod_{i=1}^{|u|-1} \bP(V = u^{(i)}) \prod_{i=0}^{|v|-1} \bP(W = v^{(i)}) \;\middle|\; \mbox{\pbox{\textwidth}{where\\$\t\vert_{\ul X} = qu$ \\ $\t\vert_{\ul Y} = qv$}} \right]\,.
  \end{IEEEeqnarray*}
}
  Now let $s\in \L_{A \to C}$ and let $t\in\sigma;\tau$ be such that $t/s$.
  By the argument in Proposition \ref{PropComposition}, there is a unique interaction sequence $S\in\sigma\|(\ul X\to\tau)\|(\Lambda\inv;\mu)$ such that $S\vert_{A,\ul{X\times Y} \to C}=t$.
  Define $\t = S\vert_{A,\ul X \to B,\ul X \to (\ul Y \to C)}$, and $\s=\t\vert_{A,B,C}$.
  Now we must have $\s\vert_{A,C}=t\vert_{A,C}=s$, so $\s\in\wit_B(s)$.

  Now consider $\t\vert_{\ul X,\ul Y}$, which has the same length as $S\vert_{\ul{X\times Y}}=t\vert_{\ul{X\times Y}}$.
  This sequence is made up of pairs of moves $qx$ for $x\in X$ or $qy$ for $y\in Y$.

  By the definition of $\mu$, we know that if the $i$-th pair of moves in $\t\vert_{\ul X,\ul Y}$ is $qx$ for $x\in X$, then the $i$-th pair of moves in $t\vert_{\ul{X\times Y}}$ is $q(x,y_0)$ for some $y_0$, and if the $i$-th pair of moves in $\t\vert_{\ul X,\ul Y}$ is $qy$ for $y\in Y$, then the $i$-th pair of mvoes in $t\vert_{\ul{X\times Y}}$ is $q(x_0,y)$ for some $x_0\in X$.
  Moreover, if $t'$ is some other sequence such that $t'/s$ and such that $t'$ differs only from $t$ in the choice of the `irrelevant' moves $x_0,y_0$, then there is some $S'\in\sigma\|(\ul X \to \tau)\|(\Lambda\inv;\mu)$ such that $S\vert_{A,\ul{X\times Y}\to C}=t'$ and $S\vert_{A,\ul X \to (\ul Y \to C)}=\t$.
  Then the combined probability of all such sequences is given by
  \[
    \left[ \prod_{i=1}^{|u|-1} \bP(V = u^{(i)}) \prod_{i=0}^{|v|-1} \bP(W = v^{(i)}) \;\middle|\; \mbox{\pbox{\textwidth}{where\\$\t\vert_{\ul X} = qu$ \\ $\t\vert_{\ul Y} = qv$}} \right]\,,
    \]
  and we can combine this with our calculations above to get the desired result.
\end{proof}

A consequence of Proposition \ref{PropCompositionProb} is that the probabilistic equivalence relation defined in \ref{DefProbabilisticEquivalenceRelation} is respected by composition of strategies.
We may therefore take the quotient of this category by the probabilistic equivalence relation to form a new category
\[
  (\G/\oppcat{(\Rv_{\Omega}^{FS})})\char`\\\approx_\bP\,.
  \]
Then next proposition shows that we do not lose anything by doing this, since the probabilistic equivalence relation gets subsumed into our intrinsic equivalence.

\begin{proposition}
  Suppose that $[\sigma],[\tau]\from A \to B$ are morphisms in the quotiented category, given by equivalence classes of strategies in $\G/\oppcat{(\Rv_\Omega^{FS})}$.
  Then $[\sigma]$ and $[\tau]$ are probabilistically intrinsically equivalent in the quotiented category if and only if $\sigma$ and $\tau$ are probabilistically intrinsically equivalent in the original category.
\end{proposition}
\begin{proof}
  If $\sigma,\tau$ are not intrinisically equivalent, then there is some $\alpha\from(A \to B) \to \bC$ such that $\bP(\Lambda\inv(\sigma);\alpha\downarrow)\ne\bP(\Lambda\inv(\tau);\alpha;\downarrow)$.
  Then we have $\bP(qa\in\Lambda\inv(\sigma);\alpha) \ne \bP(qa\in\Lambda\inv(\tau);\alpha)$, and therefore $\Lambda\inv(\sigma);\alpha\not\approx_{\bP}\Lambda\inv(\tau);\alpha$.

  Conversely, if $[\sigma],[\tau]$ are not intrinsically equivalent in the quotiented category, then there is some $[\alpha]\from (A \to B) \to \bC$ such that
  \[
    [\Lambda\inv(\sigma);\alpha] \not\approx_\bP [\Lambda\inv(\tau);\alpha]\,.
    \]
  It follows that
  \[
    \bP(qa\in\Lambda\inv(\sigma);\alpha) \ne \bP(qa\in\Lambda\inv(\tau);\alpha)\,;
    \]
  i.e., that
  \[
    \bP(\Lambda\inv(\sigma);\alpha\downarrow) \ne \bP(\Lambda\inv(\tau);\alpha\downarrow)\,,
    \]
  and so $\sigma$ and $\tau$ are not intrinsically equivalent in the original category.
\end{proof}

\section{The Probabilistic Game Semantics of Danos and Harmer}

The work in the previous section shows that it suffices to consider equivalence classes of strategies under the relation $\approx_\bP$, which suggests that we may take the quantities $\bP(s\in\sigma)$ as primitives, rather than defining them indirectly.

This is the approach taken by Danos and Harmer in \cite{DanosHarmer} -- the original game semantics for a probabilistic variant of Algol.  
Danos and Harmer define an arena as we do, but define a probabilistic strategy on an arena $A$ to be given by a function
\[
  \sigma \from \L_A^{\text{even}} \to [0,1]
  \]
such that for any even-length $s$, and any extension $sa$, we have
\[
  \sigma(s) \ge \sum_{sab\in\L_A} \sigma(sab)\,.
  \]for any even-length $s$, and any extension $sa$, we have
\[
  \sigma(s) \ge \sum_{sab\in\L_A} \sigma(sab)\,.
  \]
These quantites $\sigma(s)$ take the role of our $\bP(s\in\sigma)$, but now they are part of the \emph{definition} of the strategy $\sigma$, rather than being a calculated quantity.

\begin{proposition}
  For any \Mellies strategy $\sigma\from A \to B$ given by a random variable $V$ taking values in a set $X$ and a strategy $\tilde\sigma \from A \to (\ul X \to B)$, we get a probabilistic strategy in the sense of Danos and Harmer by setting
  \[
    \sigma(s)=\bP(s\in\sigma)\,.
    \]
\end{proposition}
\begin{proof}
  We need to check that for any odd-length legal play $sa\in \L_A$ we have
  \[
    \bP(s\in\sigma) \ge \sum_{sab\in \L_A} \bP(sab\in\sigma)\,.
    \]
  But for any $u\in\Acc_{sab}(\sigma)$, there must be some prefix $v$ of $u$ such that $v\in\Acc_s(\sigma)$.
  Then all the sequences $u$ arising in this way that have $v$ as a prefix are pairwise incomparable (since $\tilde\sigma$ is a deterministic strategy), and so their combined probability is at most the probability of $v$.
\end{proof}
Clearly, two morphisms in $\G/\oppcat{(\Rv_\Omega^{FS})}$ give rise to the same probabilistic strategy in this way if and only if they are probabilistically equivalent.

Danos and Harmer then \emph{define} the composition of two probabilistic strategies using the formula that we derived in Proposition \ref{PropCompositionProb}:

\begin{definition}
  Let $\sigma\from A \to B$, $\tau \from B \to C$ be probabilistic strategies.  
  Then their composition $\sigma;\tau$ is given by
  \[
    (\sigma;\tau)(s) = \sum_{\s\in\wit_B(s)} \sigma(\s\vert_{A,B})\tau(\s\vert_{B,C})\,.
    \]
\end{definition}
By Proposition \ref{PropCompositionProb}, the composition of strategies in the quotiented category agrees with this composition of Danos and Harmer.
Our proof above then gives an alternative proof of Full Abstraction for Danos and Harmer's probabilistic game semantics.
